---
title: "Phenotypic heterogeneity in DDG2P by allelic requirement and mutational consequence"
output: html_notebook
---

Loading DDG2P data:
```{r}
library(tidyverse)
library(readr)
ddg2p <- read_delim("DDG2P_24_9_2019.with_pLI.txt", delim = "\t")
hpo <- read_delim("table.hpo_obo_20190924.txt", delim = "\t", col_names = T)
```

First, let's take a look at the data frame:
```{r}
head(ddg2p)
head(ddg2p$phenotypes)
```

I want to separate the data in `phenotypes` column and turn the data in to a tidy format. Let's separate the data using the function `separate_rows`.
```{r}
ddg2p_tidy <- ddg2p %>% 
  separate_rows(phenotypes, sep = ';')
ddg2p_tidy
```

Let's `tally` the phenotypes.
```{r}
ddg2p_tidy %>%
  select(`gene symbol`, `allelic requirement`,  `mutation consequence`, `phenotypes`) %>%
  group_by(`gene symbol`) %>%
  drop_na() %>%
  tally(name = 'phenotypes', sort = TRUE) 
```

Looks like the gene `FLNA` has the most `phenotypes`. Now, let's add `allelic requirement` and `mutation consequence` in the function `group_by` and see the results.
```{r}
ddg2p_tidy %>%
  select(`gene symbol`, `allelic requirement`,  `mutation consequence`, `phenotypes`) %>%
  group_by(`gene symbol`,  `allelic requirement`,  `mutation consequence`) %>%
  tally(name = 'phenotypes', sort = TRUE)
```

The `allelic requirement` and `mutational consequence` does seem to have difference in the results, but `gene symbol` seems too specific for comparing.
```{r fig.height=9, fig.width=16}
p <- ddg2p_tidy %>%
  select(`allelic requirement`, `mutation consequence`, `phenotypes`) %>%
  drop_na() %>%
  group_by(`allelic requirement`, `mutation consequence`) %>%
  count(name = 'phenotypes') %>%
  ungroup() %>%
  group_by(`allelic requirement`) %>%
  mutate(allelic_sum = sum(phenotypes))

p$phenotypes <- as.numeric(p$phenotypes)
p$allelic_sum <- as.numeric(p$allelic_sum)

p
```

```{r fig.height=9, fig.width=16}
p %>% ggplot(aes(reorder(`allelic requirement`,allelic_sum), phenotypes, fill = `mutation consequence`)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  geom_text(aes(label = phenotypes), position="stack", angle = 45) +
  theme(legend.position = "top") +
  xlab("allelic requirement") +
  ggtitle("Phenotypic heterogeneity in DDG2P by allelic requirement and mutational consequence")
```

Seems much better, but the numbers are overlapping. Also, I want them to be stacked up by size, from biggest to smallest. 

```{r}
p_cumsum <- ddg2p_tidy %>%
  select(`allelic requirement`, `mutation consequence`, `phenotypes`) %>%
  drop_na() %>%
  group_by(`allelic requirement`, `mutation consequence`) %>%
  count(name = 'phenotypes') %>%
  ungroup() %>%
  group_by(`allelic requirement`) %>%
  mutate(allelic_sum = sum(phenotypes)) %>%
  arrange(-allelic_sum, `allelic requirement`) %>%
  ungroup() %>%
  group_by(allelic_sum) %>%
  arrange(-phenotypes, .by_group=TRUE) %>%
  mutate(label_ypos = cumsum(phenotypes) - 0.5*phenotypes)

p_cumsum$phenotypes <- as.numeric(p_cumsum$phenotypes)
p_cumsum$allelic_sum <- as.numeric(p_cumsum$allelic_sum)
p_cumsum$label_ypos <- as.numeric(p_cumsum$label_ypos)

p_cumsum$`mutation consequence` <- reorder(p_cumsum$`mutation consequence`, p_cumsum$phenotypes)
p_cumsum$`mutation consequence` <- factor(p_cumsum$`mutation consequence`, levels=levels(p_cumsum$`mutation consequence`))

p_cumsum
```

```{r fig.height=9, fig.width=16}
p_cumsum %>% ggplot(aes(reorder(`allelic requirement`,allelic_sum), phenotypes, fill = `mutation consequence`)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  geom_text_repel(aes(y = label_ypos, label = phenotypes))+
  theme(legend.position = "top")+
  xlab("allelic requirement") +
  ggtitle("Phenotypic heterogeneity in DDG2P by allelic requirement and mutational consequence")
```

Size difference in the data makes it difficult to tell which number is from what kind of mutation consequence. How about leaving the bar empty and coloring the text?
```{r}
install.packages("RColorBrewer")
library(RColorBrewer)
```


```{r fig.height=9, fig.width=16}
p_cumsum %>% ggplot(aes(reorder(`allelic requirement`,allelic_sum), phenotypes, col = `mutation consequence`)) +
  geom_bar(stat = "identity", fill=NA) +
  coord_flip() +
  geom_text_repel(aes(y = label_ypos, label = phenotypes))+
  theme(legend.position = "top") +
  xlab("allelic requirement") +
  ggtitle("Phenotypic heterogeneity in DDG2P by allelic requirement and mutational consequence")
```

Makes the text more readable, but hard to see the whole picture in one eye.











