---
output: html_notebook
---

# 4.4 For-loops

The formula for the sum of the series $1+2+\cdots+n$ is $n(n+1)/2$. What if we weren't sure that was the right function? How could we check? Using what we learned about functions we can create one that computes the $S_n$
```{r}
compute_s_n <- function(n){
  x <- 1:n
  sum(x)
}
```

How can we compute $S_n$ for various values of $n$, say $n = 1,...,25$? Do we write 25 lines of code calling `compute_s_n`? No, that is what for-loops are for in programming. In this case, we are performing exactly the same task over and over, and the only thing that is changing is the value of $n$. For-loops let us define the range that our variable takes (in our example $n = 1,...,10$), then change the value and evaluate expression as you _loop_.

Perhaps the simplest example of a for-loop is this useless piece of code:
```{r}
for(i in 1:5){
  print(i)
}
```

Here is the for-loop we would write for our $S_n$ example:
```{r}
m <- 25
s_n <- vector(length = m) #create an empty vector
for(n in 1:m){
  s_n[n] <- compute_s_n(n)
}
```

In each iteration $n = 1, n = 2$ etc..., we compute $S_n$ and store it in the $n$th entry of `s_n`.

Now we can create a plot to search for a pattern:
```{r}
n <- 1:m
plot(n, s_n)
```

If you noticed that it appears to be a quadratic, you are on the right track because the formula is $n(n+1)/2$, which we can confirm with a table:
```{r}
head(data.frame(s_n = s_n, formula = n*(n+1)/2))
```

We can also overlay the two results by using the function `lines` to draw a line over the previously plotted points:
```{r}
plot(n, s_n)
lines(n, n*(n+1)/2)
```

# 4.5 Vectorization and functionals

Although for-loops are an important concept to understand, in R we rarely use them. As you learn more R, you will realize that _vectorization_ is preferred over for-loops since it results in shorter and clearer code. We already saw examples in the Vector Arithmetic Section. A _vectorized_ function is a function that will apply the same operation on each of the vectors.
```{r}
x <- 1:10
sqrt(x)
y <- 1:10
x*y
```

To make this calculation, there is no need for for-loops. However, not all functions work this way. For instance, the function we just wrote, `compute_s_n`, does not work element-wise since it is expecting a scalar. This piece of code does not run the function on each entry of `n`:
```{r}
n <- 1:25
compute_s_n(n)
```

_Functionals_ are functions that help us apply the same function to each entry in a vector, matrix, data frame or list. Here we cover the functional that operates on numeric, logical and character vectors: `sapply`.

The function `sapply` permits us to perform element-wise operations on any function. Here is how it works:
```{r}
x <- 1:10
sapply(x, sqrt)
```

Each element of `x` is passed on to the function `sqrt` and the result is returned. These results are concatenated. In this case, the result is a vector of the same length as the original `x`. This implies that the for-loop above can be written as follows:
```{r}
n <- 1:25
s_n <- sapply(n, compute_s_n)
plot(n, s_n)
```

Other functionals are `apply`,`lapply`,`tapply`,`mapply`,`vapply`, and `replicate`. We mostly use `sapply`,`apply`, and `replicate` in this book, but we recommend familiarizing yourselves with the others as they can be very useful.

# 4.6 Exercises

1. What will this conditional expression return? `Not all positives`
```{r}
x <- c(1,2,-3,4)

if(all(x>0)){
  print("All Postives")
} else{
  print("Not all positives")
}
```

2. Which of the following expressions is always `FALSE` when at least one entry of a logical vector `x` is TRUE?

A. `all(x)` 
B. `any(x)`  
C. `any(!x)`  
*D. `all(!x)`*  

```{r}
x <- c(TRUE, FALSE, FALSE)
all(x)
any(x)
any(!x)
all(!x)
```
```{r}
y <- c(TRUE, TRUE)
all(y)
any(y)
any(!y)
all(!y)
```

3. The function `nchar` tells you how many characters long a character vector is.  

  Write a line of code that assigns to the object `new_names` the state abbreviation when the state name is longer than 8 characters.
```{r}
library(dslabs)
data(murders)

new_names <- ifelse(nchar(murders$state)>8, murders$abb, murders$state)
new_names
```
  
4. Create a function `sum_n` that for any given value, say $n$, computes the sum of the integers from 1 to n (inclusive). Use the function to determine the sum of integers from 1 to 5,000.
```{r}
sum_n <- function(n){
  x <- 1:n
  sum(x)
}

sum_n(5000)
```

5. Create a function `altman_plot` that takes two arguments, `x` and `y`, and plots the difference against the sum.
```{r}
altman_plot <- function(x, y){
  plot(x+y,y-x)
}
```

6. After running the code below, what is the value of `x`?
```{r}
x <- 3
my_func <- function(y){
  x <- 5
  y+5
}
x
```

7. Write a function `compute_s_n` that for any given $n$ computes the sum $S_n = 1^2 + 2^2 + 3^2 + ...n^2$. Report the value of the sum when $n = 10$.
```{r}
compute_s_n <- function(n){
  x <- 1:n
  sum(x^2)
}
compute_s_n(10)
```

8. Define an empty numerical vector `s_n` of size 25 using `s_n <- vector("numeric", 25)` and store in the results of $S_1,S_2,...S_25$ using a for-loop.
```{r}
s_n <- vector("numeric", 25)
n <- 1:25
for(i in n){
  s_n[i] <- compute_s_n(i)
}
s_n
```

9. Repeat exercise 8, but this time use `sapply`.
```{r}
s_n <- sapply(n, compute_s_n)
s_n
```

10. Repeat exercise 8, but this time use `map_dbl`.
```{r}
s_n <- map_dbl(n, compute_s_n)
s_n
```

11. Plot $S_n$ versus $n$. Use points defined by $n = 1,...,25$.
```{r}
plot(n, s_n)
```

12. Confirm that the formula for this sum is $S_n = n(n + 1)(2n + 1)/6$.
```{r}
identical(s_n, n*(n+1)*(2*n+1)/6)
```

